Computing likelihoods under the SMC for a general class of ARGs.

### 2024-07-12 paper.tex
Some text on introducing likelihood calculations

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods and Results}

First, 
we describe
in more detail what sort of information we have in estimated ARGs.
Then,
we define the SMC process in a ``backwards-in-time'' way,
rather than the usual ``along-the-genome'' way.
Note that this is a
reiteration of the original work of \citet{mcvean_approximating_2005}.

In the subsequent section we connect the two and
show how the ``backwards-in-time'' description
makes it relatively straightforward to compute the likelihood
for a more general class of estimated ARGs.
Finally, we close with a demonstration that the likelihood computation can be used
to do Bayesian inference on ancestral population sizes.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.65\textwidth]{figures/single_event.png}
    % FIXME commenting this caption out because it mucks up syntax highlighting for me.
    % \caption{
    % Graphical representation of the events associated with the edge $(p,c,x,y)$.
    % The lineage carried by $c$ spans $[x_c,y)$ until 
% a recombination event (blue dotted line) splits it
    %     into two segments. Left and right hand segments coalesce at times $t_p'$ and $t_p$ respectively,
    %     bounding the time to recombination by $(c. \min(t_p, t_{p'})$.
    %     Note that the contribution to the recombination hazard by edge $(p,c,x,y)$ only covers ``half'' of the
    %     green shaded area. The other half is contributed by edge $(p', c, x_c, x)$. 
    %     \label{fig:likelihood}
    % % }
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}

\section*{ARG likelihoods}

In one of the first approaches to probilistic ARG inference that 
regarded the ARG itself as an output rather than a latent variable,
\cite{kuhner_maximum_2000} provided an explicit formula for 
the likelihood of a realised ARG under the coalescent with
recombination:
\[
\mathbb{P}[G|N, r] = \left(\frac{1}{2N}\right)^K
r^X \exp\left(-\sum \left( \frac{1}{2N} \binom{k_i}{2} + r\ell_i\right)
t_i\right),
\]
where [explain notation].
Under this KYF formulation, the likelihood is computed by taking a 
product of the probability of each realised event, and their 
inter-event waiting times. While it is straightforward to write 
this down, computation is not so easy, and in essence requires
In essence, computing the KYF likelihood requires
``replaying'' the  state of Hudson's simulation
algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
compute the exact configuration of lineages and distribution
of ancestral material at each time point~\citep{baumdicker_efficient_2021}.

Computing the likelihood of a realised ARG under the SMC simplifies the 
bookkeeping somewhat, because we no long have the possibility of 
``trapped'' ancestral material, and each lineage can be represented 
by a single interval representing their ancestral material. Nonetheless,
we still require complete knowledge of the graph to compute the 
required probablilities.

% \section{Methods and Results}
% First, 
% we describe
% in more detail what sort of information we have in estimated ARGs.
% Then,
% we define the SMC process in a ``backwards-in-time'' way,
% rather than the usual ``along-the-genome'' way.
% Note that this is a
% reiteration of the original work of \citet{mcvean_approximating_2005}.

% In the subsequent section we connect the two and
% show how the ``backwards-in-time'' description
% makes it relatively straightforward to compute the likelihood
% for a more general class of estimated ARGs.
% Finally, we close with a demonstration that the likelihood computation can be used
% to do Bayesian inference on ancestral population sizes.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.65\textwidth]{figures/single_event.png}
    % FIXME commenting this caption out because it mucks up syntax highlighting for me.
    \caption{
    % Graphical representation of the events associated with the edge $(p,c,x,y)$.
    % The lineage carried by $c$ spans $[x_c,y)$ until 
% a recombination event (blue dotted line) splits it
    %     into two segments. Left and right hand segments coalesce at times $t_p'$ and $t_p$ respectively,
    %     bounding the time to recombination by $(c. \min(t_p, t_{p'})$.
    %     Note that the contribution to the recombination hazard by edge $(p,c,x,y)$ only covers ``half'' of the
    %     green shaded area. The other half is contributed by edge $(p', c, x_c, x)$. 
        \label{fig:likelihood}
    }
\end{figure}
